Outpacing the Lanes: Toronto's Bike Share Ridership Surges Ahead of Infrastructure Growth

### 2024-09-25 paper/paper.qmd

---
title: "Outpacing the Lanes: Toronto's Bike Share Ridership Surges Ahead of Infrastructure Growth"
author: 
  - Steven Li
thanks: "Code and data are available at: https://github.com/stevenli-uoft/Toronto_BikeShare_Development"
date: today
date-format: long
abstract: "This paper analyzes Toronto's Bike Share ridership and cycling infrastructure development from 2017 to 2023 using open data. Findings reveal a nearly fivefold increase in annual Bike Share rides, from 1.2 million to 5.7 million, outpacing the growth of cycling infrastructure. While bikeways increased from 539 in 2001 to over 1,400 by 2023, the expansion rate lags behind ridership growth. The analysis highlights a shift towards safer infrastructure, with increased focus on protected lanes. These trends underscore the need for accelerated and strategic expansion of Toronto's cycling network to meet growing demand and ensure safe, accessible cycling options for residents."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Introduction
As cities worldwide grapple with issues of congestion, pollution, and the need 
for sustainable transportation, many are turning to cycling as a potential 
solution. Toronto, Canada's largest city, has been at the forefront of this 
trend, implementing a bike-sharing program and expanding its cycling 
infrastructure to encourage more residents to adopt this eco-friendly mode of 
transport.

@@

To clean and prepare this data for analysis, we downloaded all monthly and quarterly CSV files, extracted only the required columns (trip ID and start date), and performed necessary data cleaning operations. This process involved standardizing column names across files, converting data types, and aggregating the data to calculate monthly total ridership figures. The cleaned data provides a comprehensive view of Bike Share usage from January 2017 to December 2023.

@@

@@

These two datasets were chosen as they represent the most up-to-date and accurate information available on Toronto's cycling infrastructure and Bike Share usage. Other potentially relevant datasets, such as bike lanes measured in kilometers or total bike counts, were considered but ultimately not used due to being outdated or no longer maintained.

When analyzed in tandem, these datasets provide a comprehensive picture of how Toronto's cycling infrastructure has developed in relation to the growing usage of the Bike Share program. This analysis aims to shed light on whether the city's bikeway infrastructure is keeping pace with the increasing demand for cycling as a mode of transportation in Toronto.

@@

## Bike Share Ridership Trends
The monthly ridership trend for Toronto's Bike Share program from 2017 to 2023 is illustrated in (@fig-ridership_monthly_graph). The graph shows a clear upward trajectory in ridership over the years, with pronounced seasonal fluctuations. Peak usage consistently occurs during the summer months, while winter sees a significant reduction in rides. Despite these seasonal variations, the overall trend line demonstrates steady growth in ridership.

@@

ggplot(combined_data, aes(x = year)) +
  geom_bar(aes(y = cumulative_bikeways * 2000, fill = "Bikeways"), 
           stat = "identity", alpha = 0.75) +
  geom_line(aes(y = total_rides, color = "Ridership"), size = 1) +
  scale_y_continuous(
    name = "Total Annual Bike Share Rides", 
    sec.axis = sec_axis(~ . / 2000, name = "Cumulative Number of Bikeways"),
    labels = function(x) paste0(x/1000000, "M"),
    breaks = seq(0, max(combined_data$total_rides), by = 1000000),
    minor_breaks = seq(0, max(combined_data$total_rides), by = 500000)
  ) +
  scale_x_continuous(
    breaks = seq(2017, 2023, by = 1),
    limits = c(2016.5, 2023.5),
    expand = c(0, 0)
  ) +
  scale_color_manual(values = c("Ridership" = "#1f77b4")) +
  scale_fill_manual(values = c("Bikeways" = "#2ca02c")) +
  labs(
    x = "Year",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major = element_line(color = "gray85", linewidth=0.35)
  )
```

@@

# Discussion {#sec-discussion}
The analysis of Toronto's Bike Share ridership and cycling infrastructure development from 2017 to 2023 reveals a complex interplay between infrastructure investment and cycling adoption in the city. This discussion will explore the implications of our findings, addressing the central question of whether Toronto's bikeway infrastructure is keeping pace with the growth in Bike Share ridership.

@@

1. Prioritizing Protected Infrastructure: Given the clear preference for safer cycling options, future infrastructure development should prioritize protected lanes over shared roadways where feasible.
2. Strategic Expansion: The city should consider strategically expanding the Bike Share program in tandem with new infrastructure development to ensure a balanced growth of both supply and demand.
3. Data-Driven Planning: Continued monitoring and analysis of ridership patterns and infrastructure usage can inform more targeted and effective cycling network expansion.
4. Integration with Public Transit: The success of the Bike Share program suggests opportunities for further integration with other forms of public transportation, potentially addressing last-mile connectivity issues.

## Limitations {#sec-limit}
While our study provides valuable insights into the relationship between cycling infrastructure and Bike Share usage in Toronto, it has several limitations that suggest avenues for future research:

1. Geographic Distribution: This study did not examine the spatial distribution of Bike Share usage and infrastructure. Future research could explore how infrastructure development in specific neighborhoods correlates with local Bike Share usage.
2. Quality of Infrastructure: Our analysis focused on the quantity of infrastructure rather than its quality or connectivity. Future studies could incorporate measures of infrastructure quality and network connectivity to provide a more comprehensive understanding of the cycling landscape in Toronto.
3. User Demographics: Incorporating data on user demographics could provide insights into who is benefiting from the Bike Share program and where there might be opportunities for greater inclusivity.

>>>>

---
title: "Outpacing the Lanes: Toronto's Bike Share Ridership Surges Ahead of Infrastructure Growth"
author: 
  - Steven Li
thanks: "Code and data are available at: https://github.com/stevenli-uoft/Toronto_BikeShare_Development"
date: today
date-format: long
abstract: "Toronto's Bike Share program experienced a nearly fivefold increase in annual rides from 2017 to 2023, outpacing the growth of the city's cycling infrastructure. Analysis of open data reveals that while bikeways increased from 539 in 2001 to over 1,400 by 2023, this expansion rate lags behind the surge in ridership. The mismatch between infrastructure development and Bike Share usage highlights an urgent need for accelerated and strategic expansion of Toronto's cycling network. These findings underscore the importance of data-driven urban planning to ensure safe, accessible cycling options for a growing user base."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

# Introduction
Cities worldwide, grappling with congestion and pollution, are turning to cycling as a sustainable transportation solution. Toronto, Canada's largest city, has been at the forefront of this trend, implementing a bike-sharing program and expanding its cycling infrastructure to encourage more residents to adopt this eco-friendly mode of transport.

@@

We cleaned the data by downloading all CSV files, extracting required columns (trip ID and start date), and performing necessary cleaning operations. This process involved standardizing column names across files, converting data types, and aggregating the data to calculate monthly total ridership figures. The cleaned data provides a comprehensive view of Bike Share usage from January 2017 to December 2023.

@@

Another key limitation to note about the bikeway data, is that our analysis relies on the count of bikeways rather than their total length. This approach was necessitated by data availability constraints, as information on the length of each bikeway was not provided in the dataset. Using counts instead of lengths may not provide a fully accurate representation of infrastructure growth, as it doesn't account for variations in bikeway size or capacity. For instance, a short protected lane would be given the same weight in our analysis as a long multi-use trail. This limitation should be kept in mind when interpreting the results, particularly when comparing infrastructure growth to ridership increases.

@@

These datasets, chosen for their up-to-date and comprehensive nature, offer the most accurate available information on Toronto's cycling infrastructure and Bike Share usage. When analyzed together, they provide insights into the relationship between infrastructure development and the growing demand for cycling in Toronto, allowing us to assess whether the city's bikeway expansion is keeping pace with increasing ridership. While other datasets were considered, such as bike lanes measured in kilometers or total bike counts, they were not used due to being outdated or no longer maintained.

@@

## Bike Share Ridership Trends
Toronto's Bike Share monthly ridership from 2017 to 2023 is illustrated in (@fig-ridership_monthly_graph), showing a clear upward trajectory with pronounced seasonal fluctuations. Peak usage consistently occurs during the summer months, while winter sees a significant reduction in rides. Despite these seasonal variations, the overall trend line demonstrates steady growth in ridership.

@@

ggplot(combined_data, aes(x = year)) +
  geom_bar(aes(y = cumulative_bikeways * 2000, fill = "Bikeways"), 
           stat = "identity", alpha = 0.75) +
  geom_line(aes(y = total_rides, color = "Ridership"), size = 1) +
  scale_y_continuous(
    name = "Total Annual Rides", 
    sec.axis = sec_axis(~ . / 2000, name = "Cumulative Number of Bikeways"),
    labels = function(x) paste0(x/1000000, "M"),
    breaks = seq(0, max(combined_data$total_rides), by = 1000000),
    minor_breaks = seq(0, max(combined_data$total_rides), by = 500000)
  ) +
  scale_x_continuous(
    breaks = seq(2017, 2023, by = 1),
    limits = c(2016.5, 2023.5),
    expand = c(0, 0)
  ) +
  scale_color_manual(values = c("Ridership" = "#1f77b4")) +
  scale_fill_manual(values = c("Bikeways" = "gray50")) +
  labs(
    x = "Year",
    color = NULL,
    fill = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    panel.grid.major = element_line(color = "gray85", linewidth=0.35)
  )
```

@@

# Discussion {#sec-discussion}
The analysis reveals a complex interplay between infrastructure investment and cycling adoption in Toronto from 2017 to 2023. This discussion will explore the implications of our findings, addressing the central question of whether Toronto's bikeway infrastructure is keeping pace with the growth in Bike Share ridership.

@@

1. Prioritizing Protected Infrastructure: Given the clear preference for safer cycling options, future infrastructure development should prioritize protected lanes over shared roadways where feasible.
2. Strategic Expansion: The city should consider strategically expanding the Bike Share program in tandem with new infrastructure development to ensure a balanced growth of both supply and demand.
4. Integration with Public Transit: The success of the Bike Share program suggests opportunities for further integration with other forms of public transportation, potentially addressing last-mile connectivity issues.

## Limitations and Future Research {#sec-limit}
While our study provides valuable insights into the relationship between cycling infrastructure and Bike Share usage in Toronto, it has several limitations that suggest avenues for future research:

1. **Geographic Distribution**: This study did not examine the spatial distribution of Bike Share usage and infrastructure. Future research could employ geographic information to map these distributions, revealing patterns of usage and infrastructure gaps.
2. **Quality of Infrastructure**: Our analysis focused on infrastructure quantity rather than quality or connectivity. Future studies could incorporate these measures for a more comprehensive understanding of Toronto's cycling landscape.
3. **Infrastructure Measurement**: Due to data limitations, our analysis relied on the count of bikeways rather than their length. Future studies with access to length data could offer a more precise comparison between infrastructure development and ridership growth, potentially creating a weighted index that accounts for factors such as bikeway type, length, and connectivity.
4. **Comparative Analysis**: A study comparing Toronto with cities of similar climate and urban structure could help identify best practices in balancing infrastructure development with growing cycling demand.

These research directions could significantly enhance our understanding of the complex relationship between cycling infrastructure development and ridership growth in urban environments.
